---
title: "Getting started with forrest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with forrest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  out.width  = "100%"
)
library(forrest)
```

`forrest` creates publication-ready forest plots with a single dependency
([tinyplot](https://github.com/grantmcdermott/tinyplot)) and a clean, minimal
API. It works with base R data frames, tibbles, and data.tables.

## Basic forest plot

The simplest call requires only three column names: `estimate`, `lower`, and
`upper`.

```{r basic}
dat <- data.frame(
  study    = c("Smith 2018", "Jones 2019", "Brown 2020",
               "Garcia 2021", "Lee 2022"),
  estimate = c(-0.18, 0.05, -0.12, -0.08, 0.02),
  lower    = c(-0.38, -0.12, -0.30, -0.24, -0.14),
  upper    = c(0.02, 0.22, 0.06, 0.08, 0.18)
)

forrest(
  dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "study",
  xlab     = "Mean difference (95% CI)"
)
```

## Adding a summary (pooled) estimate

Mark one or more rows as summary estimates with an `is_summary` logical column.
Those rows are drawn as filled diamonds instead of squares.

```{r summary, fig.height = 4.5}
dat$is_pooled <- c(FALSE, FALSE, FALSE, FALSE, FALSE)
dat <- rbind(dat, data.frame(
  study      = "Pooled",
  estimate   = -0.07,
  lower      = -0.16,
  upper      =  0.02,
  is_pooled  = TRUE
))

forrest(
  dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_pooled",
  xlab       = "Mean difference (95% CI)",
  title      = "Association with outcome"
)
```

## Log scale for ratio measures

When plotting odds ratios, hazard ratios, or risk ratios, set
`log_scale = TRUE` and `ref_line = 1`.

```{r log-scale, fig.height = 4.5}
or_dat <- data.frame(
  study    = c("Adams 2016", "Baker 2018", "Chen 2019",
               "Davis 2020", "Evans 2021", "Pooled"),
  estimate = c(1.45, 0.88, 1.21, 1.07, 0.95, 1.09),
  lower    = c(1.10, 0.65, 0.92, 0.80, 0.72, 0.94),
  upper    = c(1.91, 1.19, 1.59, 1.43, 1.25, 1.26),
  is_sum   = c(FALSE, FALSE, FALSE, FALSE, FALSE, TRUE)
)

forrest(
  or_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  log_scale  = TRUE,
  ref_line   = 1,
  xlab       = "Odds ratio (95% CI)"
)
```

## Subgroup headers and spacer rows

Rows whose `estimate` is `NA` are treated as visual separators. A non-empty
label is rendered in bold as a subgroup header; an empty label creates a blank
spacer line.

```{r subgroups, fig.height = 7}
sub_dat <- data.frame(
  label    = c(
    "Sex",
    "  Female", "  Male", "  Pooled",
    "",
    "Age group",
    "  18–44 years", "  45–64 years", "  65+ years", "  Pooled",
    "",
    "Overall"
  ),
  estimate = c(
    NA,  -0.10, -0.05, -0.07,
    NA,
    NA,  -0.12, -0.06, -0.02, -0.07,
    NA,
    -0.07
  ),
  lower    = c(
    NA,  -0.22, -0.18, -0.16,
    NA,
    NA,  -0.28, -0.18, -0.12, -0.16,
    NA,
    -0.16
  ),
  upper    = c(
    NA,  0.02, 0.08, 0.02,
    NA,
    NA,  0.04, 0.06, 0.08, 0.02,
    NA,
    0.02
  ),
  is_sum   = c(
    FALSE, FALSE, FALSE, TRUE,
    FALSE,
    FALSE, FALSE, FALSE, FALSE, TRUE,
    FALSE,
    TRUE
  )
)

forrest(
  sub_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "label",
  is_summary = "is_sum",
  xlab       = "Mean difference (95% CI)"
)
```

## Group coloring

Pass a `group` column to color estimates by a categorical variable. A legend
is added automatically.

```{r grouped, fig.height = 5}
grp_dat <- data.frame(
  study    = c("Smith 2018", "Jones 2019", "Brown 2020",
               "Garcia 2021", "Lee 2022", "Kim 2023"),
  estimate = c(-0.18, 0.05, -0.12, 0.22, -0.08, 0.15),
  lower    = c(-0.38, -0.12, -0.30,  0.04, -0.24, -0.01),
  upper    = c(0.02, 0.22, 0.06,  0.40, 0.08, 0.31),
  region   = c("Europe", "Europe", "Europe",
               "Americas", "Americas", "Americas")
)

forrest(
  grp_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "study",
  group    = "region",
  xlab     = "Mean difference (95% CI)"
)
```

## Adding text columns

Use the `cols` argument — a named character vector mapping display headers to
column names — to show statistics alongside the plot. This triggers a
multi-panel layout with a label panel on the left and text columns on the right.

```{r text-cols, fig.height = 5, fig.width = 9}
or_dat$or_text <- c(
  "1.45 (1.10, 1.91)",
  "0.88 (0.65, 1.19)",
  "1.21 (0.92, 1.59)",
  "1.07 (0.80, 1.43)",
  "0.95 (0.72, 1.25)",
  "1.09 (0.94, 1.26)"
)
or_dat$pval <- c("0.009", "0.372", "0.175", "0.647", "0.676", "<0.001")

forrest(
  or_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  log_scale  = TRUE,
  ref_line   = 1,
  header     = "Study",
  cols       = c("OR (95% CI)" = "or_text", "P-value" = "pval"),
  widths     = c(2, 4, 2, 1.2)
)
```

## Study weights

Pass `weight` as a column name of numeric study weights (e.g. inverse-variance
weights). Point size scales proportionally to `sqrt(weight / max(weight))`, so
larger studies appear with a bigger marker — a convention in meta-analysis
forest plots.

```{r weights, fig.height = 5}
set.seed(42)
wt_dat <- data.frame(
  study    = paste("Study", LETTERS[1:7]),
  estimate = c(-0.20, 0.04, -0.09, -0.15, 0.03, -0.08, -0.07),
  lower    = c(-0.38, -0.10, -0.24, -0.32, -0.11, -0.22, -0.16),
  upper    = c(-0.02, 0.18, 0.06, 0.02, 0.17, 0.06, 0.02),
  weight   = c(12.1, 4.5, 9.8, 7.2, 5.3, 11.0, 50.1),
  is_sum   = c(rep(FALSE, 6), TRUE)
)

forrest(
  wt_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  weight     = "weight",
  xlab       = "Mean difference (95% CI)"
)
```
