---
title: "Getting started with forrest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with forrest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
library(forrest)
```

`forrest` creates publication-ready forest plots from any data frame that
contains point estimates and confidence intervals. A single function,
`forrest()`, handles the full range of use cases — regression model results,
subgroup analyses, meta-analyses, dose-response patterns, and more.

The only hard dependency is [tinyplot](https://github.com/grantmcdermott/tinyplot).
`forrest` works with base R data frames, tibbles, and data.tables.

## Basic forest plot

The simplest call requires only three column names: `estimate`, `lower`, and
`upper`. Here we display adjusted regression coefficients from a linear model.

```{r basic}
dat <- data.frame(
  predictor = c("Age (per 10 y)", "Female sex", "BMI (per 5 kg/m²)",
                "Current smoker", "Physical activity"),
  estimate  = c(0.18, -0.42,  0.11, -0.31,  0.24),
  lower     = c(0.05, -0.61, -0.04, -0.52,  0.08),
  upper     = c(0.31, -0.23,  0.26, -0.10,  0.40)
)

forrest(
  dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "predictor",
  xlab     = "Regression coefficient (95% CI)"
)
```

## Group and subgroup headings

Rows whose `estimate` is `NA` become section headings. A non-empty label is
rendered in bold as a group header; an empty label (`""`) creates a blank spacer
line between groups.

```{r subgroups, fig.height = 5}
sub_dat <- data.frame(
  label    = c(
    "Sex",
    "  Female", "  Male",
    "",
    "Age group",
    "  Under 50", "  50 and over"
  ),
  estimate = c(NA, -0.42, -0.29, NA,  NA, -0.18, -0.51),
  lower    = c(NA, -0.61, -0.48, NA,  NA, -0.38, -0.72),
  upper    = c(NA, -0.23, -0.10, NA,  NA,  0.02, -0.30)
)

forrest(
  sub_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "label",
  xlab     = "Regression coefficient (95% CI)"
)
```

## Adding a summary (pooled) estimate

Mark one or more rows with `is_summary = TRUE` to draw them as filled diamonds
instead of squares. This is useful for pooled estimates in meta-analyses or for
overall effects after subgroup rows.

```{r summary, fig.height = 5}
sex_dat <- data.frame(
  label    = c("Female", "Male", "Overall"),
  estimate = c(-0.42, -0.29, -0.36),
  lower    = c(-0.61, -0.48, -0.50),
  upper    = c(-0.23, -0.10, -0.22),
  is_sum   = c(FALSE, FALSE, TRUE)
)

forrest(
  sex_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "label",
  is_summary = "is_sum",
  xlab       = "Regression coefficient (95% CI)",
  title      = "Association of female sex with outcome by subgroup"
)
```

## Log scale for ratio measures

When plotting odds ratios, hazard ratios, or risk ratios, set
`log_scale = TRUE` and `ref_line = 1`.

```{r log-scale, fig.height = 4.5}
or_dat <- data.frame(
  study    = c("Adams 2016", "Baker 2018", "Chen 2019",
               "Davis 2020", "Evans 2021", "Pooled"),
  estimate = c(1.45, 0.88, 1.21, 1.07, 0.95, 1.09),
  lower    = c(1.10, 0.65, 0.92, 0.80, 0.72, 0.94),
  upper    = c(1.91, 1.19, 1.59, 1.43, 1.25, 1.26),
  is_sum   = c(FALSE, FALSE, FALSE, FALSE, FALSE, TRUE)
)

forrest(
  or_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  log_scale  = TRUE,
  ref_line   = 1,
  xlab       = "Odds ratio (95% CI)"
)
```

## Group colouring

Pass a `group` column to colour estimates by a categorical variable. A legend
is added automatically using the Okabe-Ito colorblind-safe palette.

```{r grouped, fig.height = 5}
grp_dat <- data.frame(
  study    = c("Smith 2018", "Jones 2019", "Brown 2020",
               "Garcia 2021", "Lee 2022", "Kim 2023"),
  estimate = c(-0.18,  0.05, -0.12,  0.22, -0.08,  0.15),
  lower    = c(-0.38, -0.12, -0.30,  0.04, -0.24, -0.01),
  upper    = c(0.02,  0.22,  0.06,  0.40,  0.08,  0.31),
  region   = c("Europe", "Europe", "Europe",
               "Americas", "Americas", "Americas")
)

forrest(
  grp_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "study",
  group    = "region",
  xlab     = "Mean difference (95% CI)"
)
```

## Multiple estimates per row

Set `dodge = TRUE` (or a positive number) when consecutive rows share the same
`label` value. The CIs are vertically offset within each label band, and the
label is displayed once at the group centre. Combine with `group` to colour the
series.

```{r dodge, fig.height = 4, fig.width = 9}
dodge_dat <- data.frame(
  exposure = rep(
    c("Metals", "OPPs", "PFAS", "Phenols", "Phthalates"), each = 2
  ),
  period   = rep(c("Pregnancy", "Childhood"), 5),
  estimate = c(
    0.03, -0.04, -0.03,  0.04, -0.17, -0.02,  0.10, -0.02,  0.07,  0.10
  ),
  lower    = c(
    -0.25, -0.26, -0.19, -0.12, -0.40, -0.24, -0.06, -0.10, -0.15, -0.08
  ),
  upper    = c(
    0.30,  0.19,  0.13,  0.21,  0.06,  0.19,  0.26,  0.07,  0.29,  0.28
  )
)

forrest(
  dodge_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "exposure",
  group    = "period",
  dodge    = TRUE,
  header   = "Exposure class",
  ref_line = 0,
  xlab     = "Coefficient (95% CI)"
)
```

A numeric value for `dodge` sets the vertical spacing between rows within a
group directly (in y-axis units). `dodge = TRUE` uses the default of `0.25`.
Subgroup header rows (those with `NA` estimate) are always treated as singleton
groups and are not affected by dodging.

### Wide-format text columns alongside dodged CIs

By default, `cols` text values appear at each row's dodged y position,
keeping them aligned with their CI whiskers. Set `cols_by_group = TRUE` to
collapse each text column to one value per label group instead — this produces
a wide table with one row per label and one column per condition, matching the
layout commonly seen in multi-period epidemiology papers.

Populate each text column so that the value is non-empty only for the matching
condition's row, and empty (`""`) for all others; `forrest()` picks up the
right value automatically.

```{r dodge-cols, fig.height = 4, fig.width = 11}
# Same long-format data as above — just add per-condition text columns
dodge_dat$est_ci <- sprintf("%.2f (%.2f, %.2f)",
                            dodge_dat$estimate, dodge_dat$lower, dodge_dat$upper)
dodge_dat$text_preg  <- ifelse(dodge_dat$period == "Pregnancy",  dodge_dat$est_ci, "")
dodge_dat$text_child <- ifelse(dodge_dat$period == "Childhood",  dodge_dat$est_ci, "")

forrest(
  dodge_dat,
  estimate      = "estimate",
  lower         = "lower",
  upper         = "upper",
  label         = "exposure",
  group         = "period",
  dodge         = TRUE,
  cols_by_group = TRUE,
  cols          = c("Pregnancy (95% CI)" = "text_preg",
                    "Childhood (95% CI)" = "text_child"),
  widths        = c(2.5, 3.5, 2.2, 2.2),
  header        = "Exposure class",
  ref_line      = 0,
  xlab          = "Coefficient (95% CI)"
)
```

## Point shapes

Pass a `shape` column to assign different point characters per category.
Use together with `group` and `dodge` to encode two categorical dimensions
at once — for example, colour = time period and shape = sex.

```{r shape, fig.height = 4, fig.width = 9}
shape_dat <- data.frame(
  exposure = rep(c("Metals", "OPPs", "PFAS", "Phenols", "Phthalates"), each = 4),
  period   = rep(rep(c("Pregnancy", "Childhood"), each = 2), 5),
  sex      = rep(c("Female", "Male"), 10),
  estimate = c(
     0.15, -0.00, -0.09,  0.16,
    -0.07, -0.02,  0.00,  0.08,
    -0.37,  0.02, -0.08, -0.00,
     0.07,  0.11, -0.04,  0.02,
     0.16, -0.06,  0.21,  0.01
  ),
  lower    = c(
    -0.41, -0.34, -0.25, -0.19,
    -0.29, -0.25, -0.25, -0.16,
    -0.64, -0.29, -0.42, -0.28,
    -0.19, -0.13, -0.18, -0.12,
    -0.16, -0.42, -0.08, -0.30
  ),
  upper    = c(
     0.70,  0.34,  0.51,  0.34,
     0.15,  0.22,  0.25,  0.32,
    -0.10,  0.32,  0.25,  0.28,
     0.33,  0.35,  0.10,  0.16,
     0.47,  0.30,  0.51,  0.31
  )
)

forrest(
  shape_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "exposure",
  group    = "period",
  shape    = "sex",
  dodge    = TRUE,
  ref_line = 0,
  xlab     = "Coefficient (95% CI)"
)
```

The shape legend appears at `legend_shape_pos` (`"bottomright"` by default).
Set `legend_shape_pos = NULL` to suppress it.

## Adding text columns

Use the `cols` argument — a named character vector mapping display headers to
column names in `data` — to show formatted statistics alongside the plot. The
`header` argument labels the row-label panel on the left.

```{r text-cols, fig.height = 4.5, fig.width = 10}
tc_dat <- data.frame(
  predictor = c("Age (per 10 y)", "Female sex", "BMI (per 5 kg/m²)",
                "Current smoker", "Physical activity"),
  estimate  = c( 0.18, -0.42,  0.11, -0.31,  0.24),
  lower     = c( 0.05, -0.61, -0.04, -0.52,  0.08),
  upper     = c( 0.31, -0.23,  0.26, -0.10,  0.40),
  coef_ci   = c(
    " 0.18 ( 0.05,  0.31)",
    "-0.42 (-0.61, -0.23)",
    " 0.11 (-0.04,  0.26)",
    "-0.31 (-0.52, -0.10)",
    " 0.24 ( 0.08,  0.40)"
  ),
  pval = c("0.006", "<0.001", "0.148", "0.003", "0.009")
)

forrest(
  tc_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "predictor",
  header   = "Predictor",
  cols     = c("Coef (95% CI)" = "coef_ci", "P-value" = "pval"),
  widths   = c(2.8, 4, 2.5, 1.2),
  xlab     = "Regression coefficient (95% CI)"
)
```

## Alternating row stripes

Set `stripe = TRUE` for a subtle alternating row background that aids
readability, especially with many rows.

```{r stripe, fig.height = 6}
stripe_dat <- data.frame(
  label    = paste("Predictor", LETTERS[1:8]),
  estimate = c( 0.42, -0.18,  0.31, -0.07,  0.25, -0.12,  0.19, -0.34),
  lower    = c( 0.22, -0.38,  0.12, -0.24,  0.06, -0.30,  0.01, -0.52),
  upper    = c( 0.62,  0.02,  0.50,  0.10,  0.44,  0.06,  0.37, -0.16)
)

forrest(
  stripe_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "label",
  stripe   = TRUE,
  xlab     = "Regression coefficient (95% CI)"
)
```

## Study weights

Pass `weight` as a column name of numeric weights. Point size scales
proportionally to `sqrt(weight / max(weight))`, so rows with larger weights
appear with a bigger marker.

```{r weights, fig.height = 5}
wt_dat <- data.frame(
  study    = paste("Study", LETTERS[1:7]),
  estimate = c(-0.20,  0.04, -0.09, -0.15,  0.03, -0.08, -0.07),
  lower    = c(-0.38, -0.10, -0.24, -0.32, -0.11, -0.22, -0.16),
  upper    = c(-0.02,  0.18,  0.06,  0.02,  0.17,  0.06,  0.02),
  weight   = c(12.1, 4.5, 9.8, 7.2, 5.3, 11.0, 50.1),
  is_sum   = c(rep(FALSE, 6), TRUE)
)

forrest(
  wt_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  weight     = "weight",
  xlab       = "Mean difference (95% CI)"
)
```

## Saving plots

Use `save_forrest()` to export a plot to PDF, PNG, SVG, or TIFF. Pass a
zero-argument function that calls `forrest()`.

```{r save, eval = FALSE}
save_forrest(
  file   = "my_forest_plot.pdf",
  plot   = function() {
    forrest(
      dat,
      estimate = "estimate",
      lower    = "lower",
      upper    = "upper",
      label    = "predictor",
      xlab     = "Regression coefficient (95% CI)"
    )
  },
  width  = 8,
  height = 5
)
```
