---
title: "Forest plots for regression results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forest plots for regression results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 4.5,
  out.width  = "100%"
)
library(forrest)
has_broom <- requireNamespace("broom", quietly = TRUE)
```

Forest plots are not limited to meta-analyses. Any table of point estimates
with confidence intervals can be displayed as a forest plot — results from
regression models, dose-response analyses, multi-outcome studies, and more.

This vignette shows how to go from a fitted R model to a forest plot in a few
lines of code using
[`broom::tidy()`](https://broom.tidymodels.org/reference/tidy.lm.html) to
extract parameters. `broom` is not a hard dependency of `forrest`; it is used
here solely as a convenient extraction tool. You can pass any data frame with
the right columns directly to `forrest()`.

---

## Simulated data

```{r sim-data}
set.seed(42)
n <- 500

cohort <- data.frame(
  sbp      = 120 + rnorm(n, sd = 15),
  age      = runif(n, 30, 75),
  female   = rbinom(n, 1, 0.5),
  bmi      = rnorm(n, 26, 4),
  smoker   = rbinom(n, 1, 0.2),
  activity = rbinom(n, 1, 0.4)
)
# Introduce real signal
cohort$sbp <- cohort$sbp +
  0.35 * cohort$age -
  5.0  * cohort$female +
  0.4  * cohort$bmi -
  2.0  * cohort$smoker +
  2.0  * cohort$activity +
  rnorm(n, sd = 6)

# Binary outcome (hypertension: SBP ≥ 140)
cohort$htn <- as.integer(cohort$sbp >= 140)
```

---

## Multiple predictors from one model

The simplest workflow: fit a model, call `broom::tidy(conf.int = TRUE)`, and
pass the result to `forrest()`. Column names `estimate`, `conf.low`, and
`conf.high` map directly to the `estimate`, `lower`, and `upper` arguments.

```{r single-model, eval = has_broom, fig.height = 5, fig.width = 9}
fit_lm <- lm(sbp ~ age + female + bmi + smoker + activity, data = cohort)

coefs <- broom::tidy(fit_lm, conf.int = TRUE)
coefs <- coefs[coefs$term != "(Intercept)", ]

# Tidy labels
coefs$term <- c(
  "Age (per 1 year)",
  "Female sex",
  "BMI (per 1 kg/m²)",
  "Current smoker",
  "Physically active"
)

# Formatted coefficient + CI text for the right-hand column
coefs$coef_ci <- sprintf(
  "%.2f (%.2f, %.2f)",
  coefs$estimate, coefs$conf.low, coefs$conf.high
)

forrest(
  coefs,
  estimate = "estimate",
  lower    = "conf.low",
  upper    = "conf.high",
  label    = "term",
  header   = "Predictor",
  cols     = c("Coef (95% CI)" = "coef_ci"),
  widths   = c(3, 4, 2.5),
  xlab     = "Regression coefficient (95% CI)",
  stripe   = TRUE
)
```

For **logistic regression**, set `exponentiate = TRUE` in `broom::tidy()` to
get odds ratios directly, then set `log_scale = TRUE` and `ref_line = 1` in
`forrest()`:

```{r logistic, eval = has_broom, fig.height = 5}
fit_glm <- glm(htn ~ age + female + bmi + smoker + activity,
               data   = cohort,
               family = binomial)

or_coefs <- broom::tidy(fit_glm, conf.int = TRUE, exponentiate = TRUE)
or_coefs <- or_coefs[or_coefs$term != "(Intercept)", ]
or_coefs$term <- c(
  "Age (per 1 year)",
  "Female sex",
  "BMI (per 1 kg/m²)",
  "Current smoker",
  "Physically active"
)

# Add formatted OR column
or_coefs$or_text <- sprintf(
  "%.2f (%.2f, %.2f)",
  or_coefs$estimate,
  or_coefs$conf.low,
  or_coefs$conf.high
)

forrest(
  or_coefs,
  estimate   = "estimate",
  lower      = "conf.low",
  upper      = "conf.high",
  label      = "term",
  log_scale  = TRUE,
  ref_line   = 1,
  header     = "Predictor",
  cols       = c("OR (95% CI)" = "or_text"),
  widths     = c(3, 4, 2),
  xlab       = "Odds ratio (95% CI)"
)
```

---

## Same predictor across adjustment models

A common reporting pattern is to show how an association changes as
covariates are progressively added. Extract the row for the exposure of
interest from each model and combine.

```{r adjustment-models, eval = has_broom, fig.height = 4}
m1 <- glm(htn ~ age,                          data = cohort, family = binomial)
m2 <- glm(htn ~ age + female,                 data = cohort, family = binomial)
m3 <- glm(htn ~ age + female + bmi + smoker + activity, data = cohort, family = binomial)

# Extract the age row from each model
extract_age <- function(mod, label) {
  row <- broom::tidy(mod, conf.int = TRUE, exponentiate = TRUE)
  row <- row[row$term == "age", ]
  row$model <- label
  row
}

age_across_models <- rbind(
  extract_age(m1, "Model 1: unadjusted"),
  extract_age(m2, "Model 2: + sex"),
  extract_age(m3, "Model 3: + sex, BMI, smoking, activity")
)

forrest(
  age_across_models,
  estimate   = "estimate",
  lower      = "conf.low",
  upper      = "conf.high",
  label      = "model",
  log_scale  = TRUE,
  ref_line   = 1,
  xlab       = "Odds ratio for age (per 1 year, 95% CI)"
)
```

---

## Same predictor across multiple outcomes

To compare one exposure against several outcomes, loop over outcomes, extract
the exposure row from each model, and stack into a single data frame. Use
`group` to colour by outcome domain.

```{r multi-outcome, eval = has_broom, fig.height = 6}
outcomes <- list(
  list(var = "sbp",    label = "Systolic BP (mmHg)",       domain = "Continuous"),
  list(var = "htn",    label = "Hypertension (binary)",    domain = "Binary"),
  list(var = "bmi",    label = "BMI (kg/m²)",              domain = "Continuous"),
  list(var = "smoker", label = "Current smoker (binary)",  domain = "Binary")
)

rows <- lapply(outcomes, function(o) {
  if (o$domain == "Continuous") {
    fit <- lm(
      as.formula(paste(o$var, "~ activity + age + female")),
      data = cohort
    )
    r <- broom::tidy(fit, conf.int = TRUE)
  } else {
    fit <- glm(
      as.formula(paste(o$var, "~ activity + age + female")),
      data   = cohort,
      family = binomial
    )
    r <- broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE)
  }
  r <- r[r$term == "activity", ]
  r$outcome <- o$label
  r$domain  <- o$domain
  r
})
multi_out <- do.call(rbind, rows)

forrest(
  multi_out,
  estimate   = "estimate",
  lower      = "conf.low",
  upper      = "conf.high",
  label      = "outcome",
  group      = "domain",
  ref_line   = 0,
  legend_pos = "topright",
  xlab       = "Effect of physical activity (coefficient or OR, 95% CI)",
  title      = "Association of physical activity with multiple outcomes"
)
```

> **Note:** Because continuous and binary outcomes are on different scales, a
> single reference line may not be appropriate for all rows. Consider
> displaying them in separate panels, or use `ref_line = NULL` and add
> contextual annotations.

---

## Dose-response (exposure categories)

For categorical exposures (e.g., quartiles, tertiles), the reference category
has `estimate = NA` since no CI can be drawn for it. Use a text column to show
the reference label.

```{r dose-response, eval = has_broom, fig.height = 5}
# Assign quartiles of age
cohort$age_q <- cut(cohort$age, breaks = quantile(cohort$age, 0:4 / 4),
                    include.lowest = TRUE,
                    labels = c("Q1", "Q2", "Q3", "Q4"))
cohort$age_q <- relevel(cohort$age_q, ref = "Q1")

fit_q <- glm(htn ~ age_q + female + bmi + smoker + activity,
             data = cohort, family = binomial)
q_coefs <- broom::tidy(fit_q, conf.int = TRUE, exponentiate = TRUE)
q_coefs <- q_coefs[grep("^age_q", q_coefs$term), ]

# Build the full table including the reference row
q_plot <- rbind(
  data.frame(
    term      = "age_qQ1",
    estimate  = NA_real_,
    conf.low  = NA_real_,
    conf.high = NA_real_,
    or_ci     = "1.00 (ref)"
  ),
  data.frame(
    term      = q_coefs$term,
    estimate  = q_coefs$estimate,
    conf.low  = q_coefs$conf.low,
    conf.high = q_coefs$conf.high,
    or_ci     = sprintf("%.2f (%.2f, %.2f)",
                        q_coefs$estimate,
                        q_coefs$conf.low,
                        q_coefs$conf.high)
  )
)
q_plot$label  <- c("Q1 (reference)", "Q2", "Q3", "Q4")
q_plot$is_sum <- FALSE

forrest(
  q_plot,
  estimate   = "estimate",
  lower      = "conf.low",
  upper      = "conf.high",
  label      = "label",
  is_summary = "is_sum",
  log_scale  = TRUE,
  ref_line   = 1,
  header     = "Age quartile",
  cols       = c("OR (95% CI)" = "or_ci"),
  widths     = c(2.5, 4, 2),
  xlab       = "Odds ratio for hypertension (95% CI)"
)
```

---

## Marginal estimates (g-formula, IPW, etc.)

For causal methods that return marginal effects directly (parametric g-formula,
inverse probability weighting, targeted maximum likelihood estimation), the
output is typically already a data frame of estimates and confidence intervals.
Pass it directly to `forrest()` — no model extraction step is needed.

```{r gformula}
# Hypothetical marginal risk differences from a g-formula analysis
gform_dat <- data.frame(
  contrast  = c(
    "Active vs. inactive (overall)",
    "  Age < 50",
    "  Age 50–65",
    "  Age > 65"
  ),
  estimate  = c(-0.068, -0.045, -0.072, -0.089),
  lower     = c(-0.105, -0.091, -0.122, -0.148),
  upper     = c(-0.031,  0.001, -0.022, -0.030),
  is_sum    = c(TRUE, FALSE, FALSE, FALSE)
)

forrest(
  gform_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "contrast",
  is_summary = "is_sum",
  ref_line   = 0,
  xlab       = "Marginal risk difference (95% CI)",
  title      = "Effect of physical activity on hypertension risk"
)
```

---

## Saving plots

Export any plot to a file with `save_forrest()`. The device is inferred from
the file extension (`.pdf`, `.png`, `.svg`, `.tiff`).

```{r save, eval = FALSE}
save_forrest(
  file   = "regression_forest.pdf",
  plot   = function() {
    forrest(
      coefs,
      estimate = "estimate",
      lower    = "conf.low",
      upper    = "conf.high",
      label    = "term",
      stripe   = TRUE,
      xlab     = "Regression coefficient (95% CI)"
    )
  },
  width  = 9,
  height = 5
)
```
