---
title: "Getting started with forrest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with forrest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
library(forrest)
```

`forrest` creates publication-ready forest plots from any data frame that
contains point estimates and confidence intervals. A single function,
`forrest()`, handles the full range of use cases — regression model results,
subgroup analyses, meta-analyses, dose-response patterns, and more.

The only hard dependency is [tinyplot](https://github.com/grantmcdermott/tinyplot).
`forrest` works with base R data frames, tibbles, and data.tables.

## Basic forest plot

The simplest call requires only three column names: `estimate`, `lower`, and
`upper`. Here we display adjusted regression coefficients from a linear model.

```{r basic}
dat <- data.frame(
  predictor = c("Age (per 10 y)", "Female sex", "BMI (per 5 kg/m²)",
                "Current smoker", "Physical activity"),
  estimate  = c( 0.18, -0.42,  0.11, -0.31,  0.24),
  lower     = c( 0.05, -0.61, -0.04, -0.52,  0.08),
  upper     = c( 0.31, -0.23,  0.26, -0.10,  0.40)
)

forrest(
  dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "predictor",
  xlab     = "Regression coefficient (95% CI)"
)
```

## Group and subgroup headings

Rows whose `estimate` is `NA` become section headings. A non-empty label is
rendered in bold as a group header; an empty label creates a blank spacer line.
This makes it straightforward to organise results from multiple models or
exposure categories.

```{r subgroups, fig.height = 7}
multi_dat <- data.frame(
  label    = c(
    "Model 1 — unadjusted",
    "  Age (per 10 y)", "  Female sex", "  BMI (per 5 kg/m²)",
    "",
    "Model 2 — age and sex adjusted",
    "  Age (per 10 y)", "  Female sex", "  BMI (per 5 kg/m²)",
    "",
    "Model 3 — fully adjusted",
    "  Age (per 10 y)", "  Female sex", "  BMI (per 5 kg/m²)"
  ),
  estimate = c(
    NA,  0.21, -0.48,  0.14,
    NA,
    NA,  0.19, -0.44,  0.12,
    NA,
    NA,  0.18, -0.42,  0.11
  ),
  lower    = c(
    NA,  0.08, -0.68, -0.01,
    NA,
    NA,  0.06, -0.63, -0.03,
    NA,
    NA,  0.05, -0.61, -0.04
  ),
  upper    = c(
    NA,  0.34, -0.28,  0.29,
    NA,
    NA,  0.32, -0.25,  0.27,
    NA,
    NA,  0.31, -0.23,  0.26
  ),
  is_sum   = rep(FALSE, 14)
)

forrest(
  multi_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "label",
  is_summary = "is_sum",
  xlab       = "Regression coefficient (95% CI)"
)
```

## Adding a summary (pooled) estimate

Mark one or more rows with `is_summary = TRUE` to draw them as filled diamonds
instead of squares. This is useful for pooled estimates in meta-analyses or for
overall effects after subgroup rows.

```{r summary, fig.height = 5}
sex_dat <- data.frame(
  label    = c("Female", "Male", "Overall"),
  estimate = c(-0.42, -0.29, -0.36),
  lower    = c(-0.61, -0.48, -0.50),
  upper    = c(-0.23, -0.10, -0.22),
  is_sum   = c(FALSE, FALSE, TRUE)
)

forrest(
  sex_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "label",
  is_summary = "is_sum",
  xlab       = "Regression coefficient (95% CI)",
  title      = "Association of female sex with outcome by subgroup"
)
```

## Log scale for ratio measures

When plotting odds ratios, hazard ratios, or risk ratios, set
`log_scale = TRUE` and `ref_line = 1`.

```{r log-scale, fig.height = 4.5}
or_dat <- data.frame(
  study    = c("Adams 2016", "Baker 2018", "Chen 2019",
               "Davis 2020", "Evans 2021", "Pooled"),
  estimate = c(1.45, 0.88, 1.21, 1.07, 0.95, 1.09),
  lower    = c(1.10, 0.65, 0.92, 0.80, 0.72, 0.94),
  upper    = c(1.91, 1.19, 1.59, 1.43, 1.25, 1.26),
  is_sum   = c(FALSE, FALSE, FALSE, FALSE, FALSE, TRUE)
)

forrest(
  or_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  log_scale  = TRUE,
  ref_line   = 1,
  xlab       = "Odds ratio (95% CI)"
)
```

## Group colouring

Pass a `group` column to colour estimates by a categorical variable. A legend
is added automatically using the Okabe-Ito colorblind-safe palette.

```{r grouped, fig.height = 5}
grp_dat <- data.frame(
  study    = c("Smith 2018", "Jones 2019", "Brown 2020",
               "Garcia 2021", "Lee 2022", "Kim 2023"),
  estimate = c(-0.18,  0.05, -0.12,  0.22, -0.08,  0.15),
  lower    = c(-0.38, -0.12, -0.30,  0.04, -0.24, -0.01),
  upper    = c( 0.02,  0.22,  0.06,  0.40,  0.08,  0.31),
  region   = c("Europe", "Europe", "Europe",
               "Americas", "Americas", "Americas")
)

forrest(
  grp_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "study",
  group    = "region",
  xlab     = "Mean difference (95% CI)"
)
```

## Adding text columns

Use the `cols` argument — a named character vector mapping display headers to
column names in `data` — to show formatted statistics alongside the plot. This
triggers a multi-panel layout with the label panel on the left and text columns
on the right.

```{r text-cols, fig.height = 5, fig.width = 10}
or_dat$or_text <- c(
  "1.45 (1.10, 1.91)",
  "0.88 (0.65, 1.19)",
  "1.21 (0.92, 1.59)",
  "1.07 (0.80, 1.43)",
  "0.95 (0.72, 1.25)",
  "1.09 (0.94, 1.26)"
)
or_dat$pval <- c("0.009", "0.372", "0.175", "0.647", "0.676", "<0.001")

forrest(
  or_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  log_scale  = TRUE,
  ref_line   = 1,
  header     = "Study",
  cols       = c("OR (95% CI)" = "or_text", "P-value" = "pval"),
  widths     = c(2, 4, 2, 1.2)
)
```

## Alternating row stripes

Set `stripe = TRUE` for a subtle alternating row background that aids
readability, especially with many rows.

```{r stripe, fig.height = 6}
stripe_dat <- data.frame(
  label    = paste("Predictor", LETTERS[1:8]),
  estimate = c( 0.42, -0.18,  0.31, -0.07,  0.25, -0.12,  0.19, -0.34),
  lower    = c( 0.22, -0.38,  0.12, -0.24,  0.06, -0.30,  0.01, -0.52),
  upper    = c( 0.62,  0.02,  0.50,  0.10,  0.44,  0.06,  0.37, -0.16)
)

forrest(
  stripe_dat,
  estimate = "estimate",
  lower    = "lower",
  upper    = "upper",
  label    = "label",
  stripe   = TRUE,
  xlab     = "Regression coefficient (95% CI)"
)
```

## Study weights

Pass `weight` as a column name of numeric weights. Point size scales
proportionally to `sqrt(weight / max(weight))`, so rows with larger weights
appear with a bigger marker.

```{r weights, fig.height = 5}
wt_dat <- data.frame(
  study    = paste("Study", LETTERS[1:7]),
  estimate = c(-0.20,  0.04, -0.09, -0.15,  0.03, -0.08, -0.07),
  lower    = c(-0.38, -0.10, -0.24, -0.32, -0.11, -0.22, -0.16),
  upper    = c(-0.02,  0.18,  0.06,  0.02,  0.17,  0.06,  0.02),
  weight   = c(12.1, 4.5, 9.8, 7.2, 5.3, 11.0, 50.1),
  is_sum   = c(rep(FALSE, 6), TRUE)
)

forrest(
  wt_dat,
  estimate   = "estimate",
  lower      = "lower",
  upper      = "upper",
  label      = "study",
  is_summary = "is_sum",
  weight     = "weight",
  xlab       = "Mean difference (95% CI)"
)
```

## Saving plots

Use `save_forrest()` to export a plot to PDF, PNG, SVG, or TIFF. Pass a
zero-argument function that calls `forrest()`.

```{r save, eval = FALSE}
save_forrest(
  file   = "my_forest_plot.pdf",
  plot   = function() {
    forrest(
      dat,
      estimate = "estimate",
      lower    = "lower",
      upper    = "upper",
      label    = "predictor",
      xlab     = "Regression coefficient (95% CI)"
    )
  },
  width  = 8,
  height = 5
)
```
